---
title: "scRNAseq Neonatal Lung Injury 2.1"
author: Shawyon Shirazi & Nick Negretti
date: 3/13/24
output: rmarkdown::github_document
---

# Analysis of the lung epithelium

## Load libraries and helper functions
```{r, results="hide", message = FALSE, load-libraries}
source("./helper_functions/globals.R")
source("./helper_functions/libraries.R")
source("./helper_functions/cluster.R")
source("./helper_functions/colors.R")

N_WORKERS <- 12
plan("multicore", workers = N_WORKERS)
```

## Load data from merge and clean pipeline
```{r}
epi <- readRDS("./data/checkpoint/post_sct_split_epi.rds")
```

# Plot epithelial data feature counts, RNA counts, and % mitochondrial genes
```{r qc metrics violin}
epi_metric <- VlnPlot(epi,
                      c("nFeature_RNA", "nCount_RNA", "percent.mt"),
                      group.by = "condition")

ggsave("./data/checkpoint/epi/epi_metric_testing1.png",
       plot = epi_metric)
```

## UMAP and clustering part 1
```{r, fig.width = 12}
epi <- cluster_pca_umap(epi,
                        k_param = 10,
                        dims_umap = 1:15,
                        dims_neighbors = 1:15,
                        cluster_res = 1.2,
                        harmony = TRUE)

p_cluster <- DimPlot(epi, label = TRUE) +
             umap_theme() +
             theme(aspect.ratio = 1)

epi_cluster <- p_cluster +
               plot_annotation("Epithelium")
ggsave("./data/checkpoint/epi/cluster1.png",
       plot = epi_cluster)
```

```{r}
marker_genes <- c("EPCAM", "PECAM1", "PTPRC", "COL1A1", # Broad subtype
                  "AGER", "RTKN2", #AT1
                  "SFTPC", "ABCA3", "SFTPD", "LAMP3", #AT2
                  "HOPX", # AT1 and AT2
                  "KRT5", "KRT17", "TP63", # Basal
                  "FOXJ1", "TMEM190", "RSPH1", # Multiciliated
                  "SCGB3A2", "SOX4", "MGP", # RASC
                  "SCGB3A1", "MUC5B", # Goblet
                  "MKI67", "TOP2A", "CDC20", "CDK1") # Prolif and extra
epi_markers <- DotPlot(epi,
                       features = marker_genes) +
               theme(axis.text.x = element_text(angle = 45,
                                                hjust = 1))
ggsave("./data/checkpoint/epi/markers1.pdf",
       plot = epi_markers)
```

```{r metrics}
vlnplot_metrics <- VlnPlot(epi,
                           features = c("nCount_RNA",
                                        "nCount_SCT",
                                        "percent.mt",
                                        "percent.rb"),
                           group.by = "seurat_clusters")
ggsave("./data/checkpoint/epi/metrics1.png",
       plot = vlnplot_metrics,
       width = 20)
```

# Remove mesenchymal, Endothelial, epithelial, and immune doublets. 
# Remove non-biological clusters.
```{r}
epi_sub <- subset(epi,
                  idents = c(5, 18, 19, 20, 24),
                  invert = TRUE)
```

## UMAP and clustering part 2
```{r, fig.width = 12}
epi_sub <- cluster_pca_umap(epi_sub,
                            k_param = 10,
                            dims_umap = 1:15,
                            dims_neighbors = 1:15,
                            cluster_res = 1.2)

p_cluster <- DimPlot(epi_sub, label = TRUE) +
                     umap_theme() +
                     theme(aspect.ratio = 1)

epi_cluster2 <- p_cluster +
                plot_annotation("Epithelium")

ggsave("./data/checkpoint/epi/cluster2.png")
```

```{r}
marker_genes <- c("EPCAM", "PECAM1", "PTPRC", "COL1A1",
                  "AGER", "RTKN2",
                  "SFTPC", "ABCA3", "SFTPD", "LAMP3",
                  "HOPX",
                  "KRT5", "KRT17", "TP63",
                  "FOXJ1", "TMEM190", "RSPH1",
                  "SCGB3A2", "SOX4", "MGP",
                  "SCGB3A1", "MUC5B",
                  "NKX2-1", "CLDN4", "CDKN2A", "ANKRD1", "AREG", "NKD1", "LEF1",
                  "SOX9", "SLC34A2", "LCN2", "CLDN18", "SLPI", "SFTPA1",
                  "HMGA2", "AXIN2", "KRT8", "ID2", "NFKBIA", "CD44")
epi_dotplot2 <- DotPlot(epi_sub,
                        features = marker_genes) +
                theme(axis.text.x = element_text(angle = 45,
                                                 hjust = 1))
ggsave("./data/checkpoint/epi/markers2.pdf",
       width = 15)
```

```{r metrics2}
vlnplot_metrics <- VlnPlot(epi_sub,
                           features = c("nCount_RNA",
                                        "nCount_SCT",
                                        "percent.mt",
                                        "percent.rb"),
                           group.by = "seurat_clusters")
ggsave("./data/checkpoint/epi/metrics2.png",
       plot = vlnplot_metrics,
       width = 15)
``` 

```{r}
epi_sub2 <- subset(epi_sub,
                   idents = c(14),
                   invert = TRUE)
```

```{r}
N_WORKERS <- 20
plan("multicore", workers = N_WORKERS)
epi_sub2_sct <- SCTransform(epi_sub2,
                            variable.features.n = 2000,
                            batch_var = "dataset",
                            conserve.memory = TRUE,
                            vars.to.regress = c("percent.mt",
                                                "percent.rb"),
                            method = "glmGamPoi",
                            ncells = NULL)
```

## UMAP and clustering part 3
```{r soupx norm}
var_feat_backup <- VariableFeatures(epi_sub2_sct)
epi_sub2_sct <- cluster_pca_umap(epi_sub2_sct,
                                 k_param = 10,
                                 dims_umap = 1:15,
                                 dims_neighbors = 1:15,
                                 cluster_res = 1.5)
```

```{r}
p_cluster <- DimPlot(epi_sub2_sct, label = TRUE) +
             umap_theme() +
             theme(aspect.ratio = 1)

epi_cluster3 <- p_cluster +
                plot_annotation("Epithelium")
ggsave("./data/checkpoint/epi/cluster3.png")
```

```{r}
marker_genes <- c("EPCAM", "PECAM1", "PTPRC", "COL1A1",
                  "AGER", "RTKN2",
                  "SFTPC", "ABCA3", "SFTPD", "LAMP3",
                  "HOPX",
                  "KRT5", "KRT17", "TP63",
                  "FOXJ1", "TMEM190", "RSPH1",
                  "SCGB3A2", "SOX4", "MGP",
                  "SCGB3A1", "MUC5B", "MKI67",
                  "NKX2-1", "CLDN4", "CDKN2A", "ANKRD1", "AREG", "NKD1", "LEF1",
                  "SOX9", "SLC34A2", "LCN2", "CLDN18", "SLPI", "SFTPA1",
                  "HMGA2", "AXIN2", "KRT8", "ID2", "NFKBIA", "CD44")
epi_dotplot3 <- DotPlot(epi_sub2_sct,
                        features = marker_genes) +
                theme(axis.text.x = element_text(angle = 45,
                                                 hjust = 1))
ggsave("./data/checkpoint/epi/markers3.pdf",
       width = 15)
```

```{r metrics3}
vlnplot_metrics <- VlnPlot(epi_sub2_sct,
                           features = c("nCount_RNA",
                                        "nCount_SCT",
                                        "percent.mt",
                                        "percent.rb"),
                           group.by = "seurat_clusters")
ggsave("./data/checkpoint/epi/metrics3.png",
       plot = vlnplot_metrics)
FeaturePlot(epi_sub2_sct,
            c("SCGB3A2"))
```

```{r}
epi_sub3_sct <- subset(epi_sub2_sct,
                       idents = c(13, 15, 21),
                       invert = TRUE)

```

```{r}
N_WORKERS <- 20
plan("multicore", workers = N_WORKERS)
epi_sub3_sct2 <- SCTransform(epi_sub3_sct,
                             variable.features.n = 2000,
                             batch_var = "dataset",
                             conserve.memory = TRUE,
                             vars.to.regress = c("percent.mt",
                                                 "percent.rb"),
                             method = "glmGamPoi",
                             ncells = NULL)
```

## UMAP and clustering part 4
```{r}
var_feat_backup <- VariableFeatures(epi_sub3_sct2)
epi_sub3_sct2 <- cluster_pca_umap(epi_sub3_sct2,
                                  k_param = 10,
                                  dims_umap = 1:15,
                                  dims_neighbors = 1:15,
                                  cluster_res = 1.2)
```
```{r}
plot_PCA <- DimHeatmap(epi_sub3_sct2,
                       dims = 1:20,
                       nfeatures = 40,
                       fast = FALSE)
ggsave("./data/checkpoint/epi/PCA2.png",
       plot_PCA,
       height = 40)
```

```{r}
p_cluster <- DimPlot(epi_sub3_sct2, label = TRUE) +
             umap_theme() +
             theme(aspect.ratio = 1)

epi_cluster3 <- p_cluster +
                plot_annotation("Epithelium")
ggsave("./data/checkpoint/epi/cluster4.png")
```

```{r}
marker_genes <- c("EPCAM", "PECAM1", "PTPRC", "COL1A1",
                  "AGER", "RTKN2",
                  "SFTPC", "ABCA3", "SFTPD", "LAMP3",
                  "HOPX",
                  "KRT5", "KRT17", "TP63",
                  "FOXJ1", "TMEM190", "RSPH1",
                  "SCGB3A2", "SOX4", "MGP",
                  "SCGB3A1", "MUC5B", "MKI67",
                  "NKX2-1", "CLDN4", "CDKN2A", "ANKRD1", "AREG", "NKD1", "LEF1",
                  "SOX9", "SLC34A2", "LCN2", "CLDN18", "SLPI", "SFTPA1",
                  "HMGA2", "AXIN2", "KRT8", "ID2", "NFKBIA", "CD44")
epi_dotplot3 <- DotPlot(epi_sub3_sct2,
                        features = marker_genes) +
                theme(axis.text.x = element_text(angle = 45,
                                                 hjust = 1))
ggsave("./data/checkpoint/epi/markers4.pdf",
       width = 15)
```

```{r}
vlnplot_metrics <- VlnPlot(epi_sub3_sct2,
                           features = c("nCount_RNA",
                                        "nCount_SCT",
                                        "percent.mt",
                                        "percent.rb"),
                           group.by = "seurat_clusters")
ggsave("./data/checkpoint/epi/metrics4.png",
       plot = vlnplot_metrics)
```

## UMAP and clustering part 5
```{r}
var_feat_backup <- VariableFeatures(epi_sub3_sct2)

epi_sub3_sct2 <- cluster_pca_umap(epi_sub3_sct2,
                                  k_param = 10,
                                  dims_umap = 1:15,
                                  dims_neighbors = 1:15,
                                  cluster_res = 1)
```

```{r}
p_cluster <- DimPlot(epi_sub3_sct2,
                     label = TRUE) +
             umap_theme() +
             theme(aspect.ratio = 1)

epi_cluster3 <- p_cluster +
                plot_annotation("Epithelium")
ggsave("./data/checkpoint/epi/cluster5.png")
```

```{r}
marker_genes <- c("EPCAM", "PECAM1", "PTPRC", "COL1A1",
                  "AGER", "RTKN2",
                  "SFTPC", "ABCA3", "SFTPD", "LAMP3",
                  "HOPX",
                  "KRT5", "KRT17", "TP63",
                  "FOXJ1", "TMEM190", "RSPH1",
                  "SCGB3A2", "SOX4", "MGP",
                  "SCGB3A1", "MUC5B",
                  "NKX2-1", "CLDN4", "CDKN2A", "ANKRD1", "AREG", "NKD1", "LEF1",
                  "SOX9", "SLC34A2", "LCN2", "CLDN18", "SLPI", "SFTPA1",
                  "HMGA2", "AXIN2", "KRT8", "ID2", "NFKBIA", "CD44")
epi_dotplot3 <- DotPlot(epi_sub3_sct2,
                        features = marker_genes) +
                theme(axis.text.x = element_text(angle = 45,
                                                 hjust = 1))
ggsave("./data/checkpoint/epi/markers4.pdf",
       width = 15)
```

```{r metrics3}
vlnplot_metrics <- VlnPlot(epi_sub3_sct2,
                           features = c("nCount_RNA",
                                        "nCount_SCT",
                                        "percent.mt",
                                        "percent.rb"),
                           group.by = "seurat_clusters")
ggsave("./data/20240124_epi_metrics5_harmony.png",
       plot = vlnplot_metrics,
       width = 15)
```

## Save the DEgenes per cluster
```{r}
N_WORKERS <- 20
plan("multicore", workers = N_WORKERS)
filename <- "./data/checkpoint/pre_annotated_epi_for_deg.rds"
if (!file.exists(filename)) {
  markers <- parallel_find_all_markers(epi_sub3_sct2)

  names(markers) <- levels(Idents(epi_sub3_sct2))
  saveRDS(markers, filename)
} else {
  markers <- readRDS(filename)
}

wb_markers_names <- levels(Idents(epi_sub3_sct2))
wb_markers <- createWorkbook()
for (idx in seq_along(wb_markers_names)){
  addWorksheet(wb_markers,
               wb_markers_names[idx])
  writeData(wb_markers,
            wb_markers_names[idx],
            markers[[idx]],
            rowNames = TRUE)
}
saveWorkbook(wb_markers,
             file = "./data/checkpoint/epi_cluster_deg.xlsx",
             overwrite = TRUE)
```

## Annotation
```{r}
epi_relabel <- epi_sub3_sct2
epi_relabel$louvain <- as.factor(epi_sub3_sct2$sub.cluster)
epi_relabel$celltype <- as.factor(epi_sub3_sct2$sub.cluster)

idxes <- levels(epi_relabel$celltype)
labels <- rep("AT2", length(levels(epi_relabel$celltype)))
labels[idxes %in% c(4, 7, 11)] <- "RASC"
labels[idxes %in% c(9)] <- "Secretory -3A1, -3A2"
labels[idxes %in% c(1)] <- "AT1"
labels[idxes %in% c(10)] <- "Multiciliated"
labels[idxes %in% c(12)] <- "Secretory MUC5B"
labels[idxes %in% c(8)] <- "Basal"

levels(epi_relabel$celltype) <- labels
Idents(epi_relabel) <- epi_relabel$celltype
```

```{r}
DimPlot(epi_relabel,
        group.by = "celltype") +
  scale_color_manual(values = color_scanpy_default)
```

```{r}
p_cluster <- DimPlot(epi_relabel,
                     group.by = "celltype",
                     label = TRUE) +
             umap_theme() +
             scale_colour_manual(name = "Cluster",
                                 values = color_category_20) +
             theme(aspect.ratio = 1)

p_cluster + plot_annotation("Epithelium")
```

# Reorder the clusters
```{r}
epi_relabel$celltype <- ordered(as.factor(epi_relabel$celltype),
                                c("AT1",
                                  "AT2",
                                  "RASC",
                                  "Secretory -3A1, -3A2",
                                  "Secretory MUC5B",
                                  "Basal",
                                  "Multiciliated"))
```

## Save data for next step merge
```{r}
plan("multicore", workers = N_WORKERS)
filename <- "./data/checkpoint/annotated_epi.rds"
if (!file.exists(filename)) {
  saveRDS(epi_relabel,
          filename,
          compress = "bzip2")
}
```

## Save h5ad for cellxgene visualization
```{r}
writeAnnData <- function(seurat_obj, filename) {
       
  anndata::AnnData(X = t(GetAssayData(seurat_obj,
                                      slot = "data",
                                      assay = "SCT")),
    var = data.frame(gene = rownames(seurat_obj),
                     row.names = rownames(seurat_obj)),
    obs = data.frame(louvain = as.character(Idents(seurat_obj)),
                     dataset = as.character(seurat_obj$dataset),
                     seq_run = as.character(seurat_obj$orig.ident),
                     age = as.character(seurat_obj$age),
                     condition = as.character(seurat_obj$condition),
                     annotation = as.character(seurat_obj$celltype),
                     broad_condition = as.character(seurat_obj$broad_condition),
                     row.names = colnames(seurat_obj),
                     total_counts = as.numeric(seurat_obj$nCount_SCT),
                     pct_mito = as.numeric(seurat_obj$percent.mt),
                     pct_ribo = as.numeric(seurat_obj$percent.rb)),
    obsm = list(X_umap = matrix(Embeddings(seurat_obj,
                                           reduction = "umap"),
                                ncol = 2)),
    raw = anndata::AnnData(X = t(GetAssayData(seurat_obj,
                                              slot = "counts",
                                              assay = "RNA")),
                           var = data.frame(gene = rownames(seurat_obj@assays$RNA),
                                            row.names = rownames(seurat_obj@assays$RNA)))
  )$write_h5ad(filename,
               compression = "gzip")

}

filename <- "./data/checkpoint/epi_only.h5ad"
if (!file.exists(filename)) { # Create if doesn"t exist
  writeAnnData(epi_relabel,
               filename)
}
```